---
title: "Chapter 1 - Introduction to Data"
author: 'Amit Kapoor'
date: '8/29/2019'
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Smoking habits of UK residents**. (1.10, p. 20) A survey was conducted to study the smoking habits of UK residents. Below is a data matrix displaying a portion of the data collected in this survey. Note that “£” stands for British Pounds Sterling, "cig" stands for cigarettes, and "N/A" refers to a missing component of the data.

\begin{center}
\scriptsize{
\begin{tabular}{rccccccc}
\hline
	& sex 	 & age 	& marital 	& grossIncome 					     & smoke & amtWeekends	& amtWeekdays \\ 
\hline
1 	& Female & 42 	& Single 	& Under $\pounds$2,600 			     & Yes 	 & 12 cig/day   & 12 cig/day \\ 
2 	& Male	 & 44	& Single 	& $\pounds$10,400 to $\pounds$15,600 & No	 & N/A 			& N/A \\ 
3 	& Male 	 & 53 	& Married   & Above $\pounds$36,400 		     & Yes 	 & 6 cig/day 	& 6 cig/day \\ 
\vdots & \vdots & \vdots & \vdots & \vdots 				             & \vdots & \vdots 	    & \vdots \\ 
1691 & Male  & 40   & Single 	& $\pounds$2,600 to $\pounds$5,200   & Yes 	 & 8 cig/day 	& 8 cig/day \\   
\hline
\end{tabular}
}
\end{center}


(a) What does each row of the data matrix represent?
**Each row of a data matrix represents a unique case (observation unit).**

(b) How many participants were included in the survey?
**1691**

(c) Indicate whether each variable in the study is numerical or categorical. If numerical, identify as continuous or discrete. If categorical, indicate if the variable is ordinal.
* **Sex --> Categorical - nominal**
* **Age --> Numerical - discrete**
* **Marital --> Categorical - nominal**
* **grossIncome --> Categorical - ordinal**
* **smoke --> Categorical - nominal**
* **amtWeekends --> Numerical - discrete**
* **amtWeekdays --> Numerical - discrete**

```{r, eval=FALSE, echo=FALSE}
# NOTE: It is not absolutely necessary to load the data to answer these 
# questions, but is here for further exploration. 
if(!file.exists('11263-Smoking_tcm86-13253.xls')) {
	download.file('https://github.com/jbryer/DATA606Fall2019/blob/master/course_data/11263-Smoking_tcm86-13253.xls?raw=true', 
				  destfile = '11263-Smoking_tcm86-13253.xls')
}
smoking <- readxl::read_excel('11263-Smoking_tcm86-13253.xls')
```





--------------------------------------------------------------------------------

\clearpage

**Cheaters, scope of inference**. (1.14, p. 29) Exercise 1.5 introduces a study where researchers studying the relationship between honesty, age, and self-control conducted an experiment on 160 children between the ages of 5 and 15[^1]. The researchers asked each child to toss a fair coin in private and to record the outcome (white or black) on a paper sheet, and said they would only reward children who report white. Half the students were explicitly told not to cheat and the others were not given any explicit instructions. Differences were observed in the cheating rates in the instruction and no instruction groups, as well as some differences across children’s characteristics within each group.

(a) Identify the population of interest and the sample in this study.
**children between the ages of 5 and 15, sample - 160**

(b) Comment on whether or not the results of the study can be generalized to the population, and
if the findings of the study can be used to establish causal relationships.
**Since this is an experiment conducted and not just an observational study, it can be used to establish causal relationships. My concern is to generalize the study to the population given the confounding variable like grouping more younger or older children than others. The study can only be generalized to children similar to the ones studied in this experiment.**

[^1]: Alessandro Bucciol and Marco Piovesan. “Luck or cheating? A field experiment on honesty with children”. In: Journal of Economic Psychology 32.1 (2011), pp. 73–78. Available at https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1307694



--------------------------------------------------------------------------------

\clearpage

**Reading the paper**. (1.28, p. 31) Below are excerpts from two articles published in the NY Times:

(a) An article titled Risks: Smokers Found More Prone to Dementia states the following:

"Researchers analyzed data from 23,123 health plan members who participated in a voluntary exam and health behavior survey from 1978 to 1985, when they were 50-60 years old. 23 years later, about 25% of the group had dementia, including 1,136 with Alzheimer's disease and 416 with vascular dementia. After adjusting for other factors, the researchers concluded that pack-a- day smokers were 37% more likely than nonsmokers to develop dementia, and the risks went up with increased smoking; 44% for one to two packs a day; and twice the risk for more than two packs."

Based on this study, can we conclude that smoking causes dementia later in life? Explain your reasoning.

**No, we can’t conclude that smoking causes dementia later in life since the study is based out of observation and not  experiment. The other factor to consider here is that participants volunteered and not randomly selected.**

(b) Another article titled The School Bully Is Sleepy states the following:

"The University of Michigan study, collected survey data from parents on each child's sleep habits and asked both parents and teachers to assess behavioral concerns. About a third of the students studied were identified by parents or teachers as having problems with disruptive behavior or bullying. The researchers found that children who had behavioral issues and those who were identified as bullies were twice as likely to have shown symptoms of sleep disorders."

A friend of yours who read the article says, "The study shows that sleep disorders lead to bullying in school children." Is this statement justified? If not, how best can you describe the conclusion that can be drawn from this study?

**The statement cannot be justified given that it is again on an observation data. We can only conclude that sleeping disorder can be correlated to bullying. There can be other reasons for sleeping disorder like lack of sleep which can be considered as confounding variable.**



--------------------------------------------------------------------------------

\clearpage

**Exercise and mental health.** (1.34, p. 35) A researcher is interested in the effects of exercise on mental health and he proposes the following study: Use stratified random sampling to ensure rep- resentative proportions of 18-30, 31-40 and 41-55 year olds from the population. Next, randomly assign half the subjects from each age group to exercise twice a week, and instruct the rest not to exercise. Conduct a mental health exam at the beginning and at the end of the study, and compare the results.

(a) What type of study is this? 
**This study is an experiment since here the researcher assigns treatments to cases by asking a group to exercise and the other not to exercise. Here the explanatory variable is ‘exercising or not’ and response variable is result of ‘mental health exam’.**
(b) What are the treatment and control groups in this study?
**Treatment group is the group asked to exercise twice a week and control group is one asked not to exercise.**
(c) Does this study make use of blocking? If so, what is the blocking variable?
**Yes it does use blocking based on age groups (18-30, 31-40 and 41-55).**
(d) Does this study make use of blinding?
**No, it doesn’t use blinding.**
(e) Comment on whether or not the results of the study can be used to establish a causal rela- tionship between exercise and mental health, and indicate whether or not the conclusions can be generalized to the population at large.
**It is a randomized experiment and not an observation study so it can be used to establish a causal relationship between exercise and mental health. The sample size would be a factor to determine if the conclusion can be generalized to large population.**
(f) Suppose you are given the task of determining if this proposed study should get funding. Would you have any reservations about the study proposal?
**The study is good but I do have suggestions:**
* **Make sample size large enough to make it practical.**
* **Include the entire population and not just certain age groups.**
* **Make experiment to consider blinding.**



