---
title: "Chapter 2 - Summarizing Data"
author: "Amit Kapoor"
date: "9/7/2019"
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Stats scores**. (2.33, p. 78) Below are the final exam scores of twenty introductory statistics students. 

57, 66, 69, 71, 72, 73, 74, 77, 78, 78, 79, 79, 81, 81, 82, 83, 83, 88, 89, 94

Create a box plot of the distribution of these scores. The five number summary provided below may be useful.

\begin{center}
\renewcommand\arraystretch{1.5}
\begin{tabular}{ccccc}
Min & Q1    & Q2 (Median)   & Q3    & Max \\
\hline
57  & 72.5  & 78.5          & 82.5  & 94 \\
\end{tabular}
\end{center}

```{r scores-boxplot, echo = FALSE}
scores <- c(57, 66, 69, 71, 72, 73, 74, 77, 78, 78, 79, 79, 81, 81, 82, 83, 83, 88, 89, 94)
boxplot(scores)
```




--------------------------------------------------------------------------------

\clearpage

**Mix-and-match**. (2.10, p. 57) Describe the distribution in the histograms below and match them to the box plots.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
set.seed(7365)
sym = rnorm(1000, mean = 60, sd = 3)
uni = runif(1000, min = 0, max = 100)
rs = rgamma(1000, shape = 3, rate = 2)
par(mar=c(4, 3.6, 0, 0), las = 1, mgp = c(2.7, 0.7, 0), 
    mfrow = c(1,6), 
    cex.lab = 1.5, cex.axis = 1.5)
histPlot(sym, xlab = "(a)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(50,70,10))
histPlot(uni, xlab = "(b)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(0,100,50))
histPlot(rs, xlab = "(c)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(0,6,2))
boxPlot(rs, xlab = "(1)", ylab = "", col = COL[1,3])
boxPlot(sym, xlab = "(2)", ylab = "", col = COL[1,3])
boxPlot(uni, xlab = "(3)", ylab = "", col = COL[1,3])
```


* **The boxplot that matches histogram (a) is (2). The distribution of this data is unimodal appears to be normally distributed.**
* **The boxplot that matches histogram (b) is (3). The distribution of this data is multimodal and could be symmetric but doesn't look normally distributed.**
* **The boxplot that matches histogram (c) is (1). The distribution of this data is right skewed and appears unimodal.**


--------------------------------------------------------------------------------

\clearpage

**Distributions and appropriate statistics, Part II**. (2.16, p. 59) For each of the following, state whether you expect the distribution to be symmetric, right skewed, or left skewed. Also specify whether the mean or median would best represent a typical observation in the data, and whether the variability of observations would be best represented using the standard deviation or IQR. Explain your reasoning.

(a) Housing prices in a country where 25% of the houses cost below $350,000, 50% of the houses cost below $450,000, 75% of the houses cost below $1,000,000 and there are a meaningful number of houses that cost more than $6,000,000.

**Right Skewed since there are a meaningful number of houses that cost more than $6,000,000. Median and IQR would be the best mesurements.**

(b) Housing prices in a country where 25% of the houses cost below $300,000, 50% of the houses cost below $600,000, 75% of the houses cost below $900,000 and very few houses that cost more than $1,200,000.

**Symmetric and uniformly distributed since very few houses that cost more than $1,200,000 will not skew. Mean and Standard deviation would be the best mesurements.**

(c) Number of alcoholic drinks consumed by college students in a given week. Assume that most of these students don’t drink since they are under 21 years old, and only a few drink excessively.

**Right Skewed since there isn’t a significant number of excessive drinkers. Median and IQR would be the best mesurements.**

(d) Annual salaries of the employees at a Fortune 500 company where only a few high level
executives earn much higher salaries than the all other employees.

**Left Skewed. Mean and IQR would be the best measurements.**



--------------------------------------------------------------------------------

\clearpage

**Heart transplants.** (2.26, p. 76) The Stanford University Heart Transplant Study was conducted to determine whether an  experimental heart transplant program increased lifespan. Each patient entering the program was designated an official heart transplant candidate, meaning that he was gravely ill and would most likely benefit from a new heart. Some patients got a transplant and some did not. The variable *transplant* indicates which group the patients were in; patients in the treatment group got a transplant and those in the control group did not. Of the 34 patients in the control group, 30 died. Of the 69 people in the treatment group, 45 died. Another variable called *survived* was used to indicate whether or not the patient was alive at the end of the study.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=3}
library(openintro)
data(heartTr)
# mosaic plot -------------------------------------------------------
par(mar = c(0, 0, 0, 0), las = 1, mgp = c(2.7, 0.9, 0))
mosaicplot(transplant ~ survived, data = heartTr, 
           main = "", xlab = "", ylab = "", color = COL[1],
           cex.axis = 1.5)

# box plot ----------------------------------------------------------
par(mar = c(2, 4.8, 0, 0), las = 1, mgp = c(3.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
boxPlot(heartTr$survtime, fact = heartTr$transplant, 
        ylab = "Survival Time (days)", col = COL[1,2])
```

(a) Based on the mosaic plot, is survival independent of whether or not the 
patient got a transplant? Explain your reasoning.

**survival is dependent upon whether or not patient got a transplant because in the mosaic plot its clearly visible that the survival of patients who got treatment is more then who didnt get.**

(b) What do the box plots below suggest about the efficacy (effectiveness) of the heart transplant treatment.

**boxplot shows that the patients who received the treatment has more survival time (days).**

(c) What proportion of patients in the treatment group and what proportion of 
patients in the control group died?

```{r head-heartTr, eval=TRUE}
head(heartTr)
```

```{r compute-dead, eval=TRUE}
trtmnt_patient <- nrow(subset(heartTr, transplant == "treatment"))
trtmnt_patient_dead <- nrow(subset(heartTr, transplant == "treatment" & survived =="dead"))
ctrl_patient <- nrow(subset(heartTr, transplant == "control"))
ctrl_patient_dead <- nrow(subset(heartTr, transplant == "control" & survived =="dead"))
```

```{r treatment-dead-ratio, eval=TRUE}
trtmnt_patient_dead/trtmnt_patient
```

**65.22% patients in treatment group died.**

```{r control-dead-ratio, eval=TRUE}
ctrl_patient_dead/ctrl_patient
```

**88.24% patients in control group died.**

(d) One approach for investigating whether or not the treatment is effective 
is to use a randomization technique.

i. What are the claims being tested?

**The claim being tested is whether or not a heart transplant increases the lifespan of patients.**

ii. The paragraph below describes the set up for such approach, if we were 
to do it without using statistical software. Fill in the blanks with a number 
or phrase, whichever is appropriate.

```{r heartTr-alive, eval=TRUE}
nrow(subset(heartTr, survived == "alive"))
```


```{r heartTr-dead, eval=TRUE}
nrow(subset(heartTr, survived == "dead"))
```

```{r trtmnt-patient, eval=TRUE}
trtmnt_patient
```

```{r ctrl-patient, eval=TRUE}
ctrl_patient
```

```{r diff-ratio, eval=TRUE}
0.8823529 - 0.6521739
```


> We write *alive* on **28** cards representing patients who were 
alive at the end of the study, and *dead* on **75** cards 
representing patients who were not. Then, we shuffle these cards and split them 
into two groups: one group of size **69** representing treatment, and 
another group of size **34** representing control. We calculate the 
difference between the proportion of *dead* cards in the treatment and 
control groups (treatment - control) and record this value. We repeat this 100 
times to build a distribution centered at **0**. Lastly, we calculate 
the fraction of simulations where the simulated differences in proportions are 
**0.2302**. If this fraction is low, we conclude that it is unlikely to 
have observed such an outcome by chance and that the null hypothesis should 
be rejected in favor of the alternative.

iii. What do the simulation results shown below suggest about the effectiveness 
of the transplant program?

```{r fig.height=3, fig.show="hold", message=FALSE, warning=FALSE, include=FALSE, out.width="50%", results='hide'}
# randomization -----------------------------------------------------
diffs <- DATA606::inference(heartTr$survived, heartTr$transplant, 
                  success = "dead", order = c("treatment","control"), 
                  est = "proportion", type = "ht", method = "simulation", 
                  nsim = 100, null = 0, alternative = "twosided", simdist = TRUE,
                  seed = 95632)
```

**simulation results suggest that transplant program is effective since the difference is centered around 0 and data shows strong evidence of higher survival with treatment. Thus I reject the null hypothesis.**


\begin{center}
```{r, results='hide', echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="75%", fig.height=3}
# plot randomization distribution -----------------------------------
par(mar = c(3.6, 2.2, 1, 1), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.axis = 1.5, cex.lab = 1.5)
values  <- table(diffs)
plot(diffs, type = "n", xlim = c(-0.25, 0.25), 
     xlab = "simulated differences in proportions", 
     ylab = "", axes = FALSE, ylim = c(0, max(values)))
axis(1, at = seq(-0.25, 0.25, 0.05), 
     labels = c(-0.25, NA,-0.15, NA,-0.05, NA, 0.05, NA, 0.15, NA, 0.25))
for(i in 1:length(diffs)){
  x   <- diffs[i]
  rec <- sum(diffs == x)
  points(rep(x, rec), 1:rec, pch = 20, cex = 0.8, col = COL[1])
}
```
\end{center}





